---
title: "Getting started with cfaTools"
author: "Damian W Betebenner"
date: "`r toOrdinal::toOrdinalDate()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with cfaTools}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r include = FALSE}
library(cfaTools)
is_html_output = function() {
  knitr::opts_knit$get("rmarkdown.pandoc.to")=="html"
}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "",
    prompt=TRUE,
    fig.dpi = 96
			)
if (is_html_output()) {
    options(width=1000)
}
```

# Introduction

The R package **cfaTools** contains data analysis tools for use in examining large scale student educational assessment data. The package is under active development by staff at the [Center for Assessment](https://www.nciea.org).

# Use

After [installing](../index.html#installation) the package from either CRAN
or GitHub, there are many possible analyses of educational assessments that can be conducted using functions in the package.


## Examples

The first argument, `n`, is the cardinal number to be converted to an ordinal number:

```{r eval=FALSE}
cfaGaps(gap_data=sgpData_LONG[YEAR=="2018_2019", c("SCALE_SCORE", "ETHNICITY"), with=FALSE],
        gap_groups=c("African American", "White"),
	output_format="PLOTLY")
```


# Contributions & Requests

If you have contribution (for example, another language) or a feature request for the **cfaTools** package, don't hesitate to write or set up an [issue on GitHub](https://github.com/CenterForAssessment/cfaTools/issues).
