---
title: "Simulating Missing Data with amputeScaleScore"
author: "Adam Van Iwaarden"
date: "June 10, 2021"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating Missing Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
header-includes:
- \usepackage{multirow}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In preparation for the analysis of student assessment data in 2021, and given the
assumption of reduced test participation in many states due to the ongoing COVID-19
pandemic, simulations of missing data that produce varying degrees and patterns
of missingness were required.  The `amputeScaleScore` function allows for customized
simulation of data missing completely at random (MCAR), missing at random (MAR)
and missing not at random (MNAR). These patterns can be specified to depend on a
variety of student and institutional level variables.

This vignette provides a brief introduction of how to use the `amputeScaleScore`
function to simulate, explore and analyze missing data patterns using the
`sgpData_LONG_COVID` data from the the [SGPData](https://github.com/CenterForAssessment/SGPdata)
package. This simulated data conforms to the data naming and structure that is
[required for data](https://centerforassessment.github.io/SGPdata/reference/index.html#section-data-sets-for-higher-level-sgp-functions)
analysis using the [SGP package](https://github.com/CenterForAssessment/SGP).


# Function Arguments

The following provides an explanation of the arguments a user can provide in the
`amputeScaleScore` function.

```{r arguments}
  require(cfaTools)
  args(amputeScaleScore)
```
* `ampute.data`
  - The complete dataset in which to create missing values
  - Here the COVID impacted dataset from SGPdata: `SGPdata::sgpData_LONG_COVID`

* `additional.data = NULL`
  - The function will return only data that is required for a SGP analysis (based
    on `growth.config`) or a status analysis (based on `status.config`). This allows
    for the addition of more data (e.g. 2019 grades not used as priors).

* `compact.results = FALSE`
  - By default (`FALSE`), the function will return a list of longitudinal datasets
    with the current (amputed) and prior (unchanged) student records.  This is helpful
    for diagnostics and ease of use, but also produces redundant prior data. Setting
    this argument to `TRUE` returns a single `data.table` object with a TRUE/FALSE
    indicator column added for each requested amputation. This flag can be used
    to make the `SCALE_SCORE` (and/or other variables) `NA` in subsequent use cases.

* `growth.config = NULL`
  - An elongated SGP config script. This needs to have an entry for each
    grade/content_area/year cohort that will be analyzed in subsequent simulations.

* `status.config = NULL`
  - An elongated SGP config script. This needs to have an entry for each
    grade/content_area/year cohort that will be analyzed in subsequent simulations.
  - Unlike a `growth.config` entry, `status.config` entries use data from the same
    grade, but from a prior year (i.e. not individual variables). For example you
    might predict missing 2021 3rd grade ELA scores based on 2019 3rd grade school
    mean scale score, FRL status, etc.

* `default.vars = c("CONTENT_AREA", "GRADE", "SCALE_SCORE", "ACHIEVEMENT_LEVEL")`
  - variables that will be used in extracting cohort records (subject and grade) and
    other variables from the data that the user may want returned in the amputed data.

* `demographics = NULL`
  - Demographic (factor/character) variables to use and/or return in the final dataset.
  - For example:
    `c("FREE_REDUCED_LUNCH_STATUS", "ELL_STATUS", "IEP_STATUS", "ETHNICITY", "GENDER")`

* `institutions = NULL`
  - Institution IDs that will be used/or returned in the amputed data.
  - For example:
    `c("SCHOOL_NUMBER", "DISTRICT_NUMBER")`

* `ampute.vars = NULL`
  - Intersection of `default.vars`, `demographics` and `institutions` that will
    be used in the construction of the weighted scores that define the probability
    of being missing. Any institution included will be used to construct institution
    level means of other student level `ampute.vars.`
  - For example, with
    `c("SCHOOL_NUMBER", "SCALE_SCORE", "FREE_REDUCED_LUNCH_STATUS")`, student
    level scores and demographics will be used along with their associated school
    level mean scores and proportion FRL (4 total factors).
  - The default (`NULL`) means that no factors are considered, creating a
    "missing completely at random" (MCAR) missing data pattern.

* `ampute.var.weights = NULL`
  - Relative weights assigned to the `ampute.vars`.  Default is `NULL` meaning
    the weighted sum scores will be calculated with equal weight (=1). A named
    list can be provided with the desired relative weights. For example,
    `list(SCALE_SCORE=3, FREE_REDUCED_LUNCH_STATUS=2, SCHOOL_NUMBER=1)` will weight
    a student's scale scores by a factor of 3 and FRL by 2, with all (any other
    `ampute.vars`) remaining at the default of 1.
  - School level aggregates are included in the example above.  Note that differential
    weights for institutions should be placed at the end of the list. If institution
    IDs (e.g., SCHOOL_NUMBER) are omitted from the list, the aggregates will be given
    the same weight as the associated student level variable. In the given example,
    `SCALE_SCORE` and school mean scale score would be given a relative weight of 3.
  - This argument is ignored when `ampute.vars = NULL`.

* `reverse.weight = "SCALE_SCORE"`
  - The current default for `ampute.args$type` is `"RIGHT"`, which means that
    students with high weighted scores have the highest probability for amputation.
    This makes sense for high % FRL schools, but not for high achieving students
    and/or students in high achieving schools. This function inverses the variable(s)
    individual (and institutional mean) value(s) so that higher weight is given
    to lower scores/means.
  - This argument is ignored when `ampute.vars = NULL`.

* `ampute.args = list(prop=0.3, type="RIGHT")`
  - variables to be used in the `mice:ampute.continuous` function.  Currently only
    `prop` and `type` can be modified.  See `?mice::ampute.continuous` and
    `?mice::ampute` for more information.
  - The `prop` argument is inexact and has required some modification. Although
    performance has been improved, it is still imprecise, particularly for values
    away from 0.5 (50% missing).  Also, the max missingness is 85%, and for that
    the user needs to set `prop=0.95` or greater.
  - Note that `prop` gives a total proportion missing, which accounts for missingness
    already included in the data due to students with incomplete longitudinal data
    histories.  For example, if a cohort starts with 5% students missing due to
    incomplete histories, an additional 25% will be made missing to get to the 30%
    (default) missingness.
  - This last point has been dealt with in some regards with the next argument,
    which removes these cases first.

* `complete.cases.only = TRUE`
  - Should cases without the most recent prior and current score be removed?
  - This removes students with partial longitudinal histories from the most recent
    prior (e.g., 2019) to the current year (e.g., 2021), producing a "complete"
    dataset that is easier to interpret and use in subsequent imputation simulations.

* `partial.fill = TRUE`
  - Should an attempt be made to fill in some of the demographic and institution
    ID information based on students previous values?
  - Part of the process of the `amputeScaleScore` function is to take the long data
    (`ampute.data`) and then first widen and then re-lengthen the data, which
    creates holes for students with incomplete longitudinal records.  This part
    of the function fills in these holes for students with existing missing data.

* `invalidate.repeater.dups = TRUE`
  - Students who repeat a grade will get missing data rows inserted for the grade
    that they "should" be in in 2021. This leads to duplicated cases that can lead
    to problems in the SGP analyses. This argument returns those cases with the
    `VALID_CASE` variable set to `"INVALID_CASE"`.

* `seed = 4224L`
  - A random seed set for the amputation process to allow for replication of
    results, or for alternative results using the same code.

* `M = 10`
  - The number of amputed datasets to return.  The default is 10.

Depending on the amputation process specified, the function returns either a list
(default) of `M` amputed datasets or a single data set with `M` columns of missing
record indicators. The datasets will exclude data for students not used in any of
the specified `growth.config` or `status.config` cohorts, unless the `additional.data`
argument has been included.


## Example workflow of creating missing data

The sgpData_LONG_COVID object contains simulated data from 2016 to 2023. The data
follows the same format as the example [long data](https://sgp.io/articles/SGP_Data_Preparation.html#long-data-format-sgpdata_long)
from the SGP package. The dataset is 7 years of annual assessment data in two content
areas (ELA and Mathematics) and is missing 2020 data to help users model COVID related
interruptions to student status and growth. The data comes with a "built in" impact
in 2021 related to the pandemic (although an unperturbed version - `"SCALE_SCORE_without_COVID_IMPACT"`
is also available so that different impact scenarios can be modeled).

```{r packages}
  ###   Load packages
  require(data.table)
```

We will use a copy of the ELA data from the SGPdata::sgpData_LONG_COVID dataset.
We will also add duplicate `SCALE_SCORE` and `ACHIEVEMENT_LEVEL` variables to compare
with what has been amputated (named `SCALE_SCORE_COMPLETE` and `ACH_LEV_COMPLETE`).

```{r data_to_ampute}
  data_to_ampute <-
      data.table::copy(SGPdata::sgpData_LONG_COVID)[CONTENT_AREA == "ELA",
                       SCALE_SCORE_COMPLETE := SCALE_SCORE][,
                       ACH_LEV_COMPLETE := ACHIEVEMENT_LEVEL]
```

Next we create the required configuration scripts. Here is an example of the "status
only" grade configurations.  That is, grades 3 and 4 will not have prior scores
from 2020 available to use as either a factor in score amputation/imputation or
in student growth analyses.

```{r status_config}
  ###   ELA 2021 status configurations for amputeScaleScore

  status_config_2021 <- list(
    ELA.2021 = list(
      sgp.content.areas = rep("ELA", 2),
      sgp.panel.years = c("2019", "2021"),
      sgp.grade.sequences = c("3", "3")),
    ELA.2021 = list(
      sgp.content.areas = rep("ELA", 2),
      sgp.panel.years = c("2019", "2021"),
      sgp.grade.sequences = c("4", "4"))
  )
```

```{r growth_config}
  ###   ELA 2021 growth configurations for amputeScaleScore

  growth_config_2021 <- list(
    ELA.2021 = list(
      sgp.content.areas = rep("ELA", 2),
      sgp.panel.years = c("2019", "2021"),
      sgp.grade.sequences = c("3", "5")),
    ELA.2021 = list(
      sgp.content.areas = rep("ELA", 3),
      sgp.panel.years = c("2018", "2019", "2021"),
      sgp.grade.sequences = c("3", "4", "6")),
    ELA.2021 = list(
      sgp.content.areas = rep("ELA", 3),
      sgp.panel.years = c("2018", "2019", "2021"),
      sgp.grade.sequences = c("4", "5", "7")),
    ELA.2021 = list(
      sgp.content.areas = rep("ELA", 3),
      sgp.panel.years = c("2018", "2019", "2021"),
      sgp.grade.sequences = c("5", "6", "8"))
  )
```

###  A quick example run

To begin with we will run a quick test using the package defaults along with the
augmented data and config scripts we just read in. The following script creates
a single amputed dataset (which takes about as long as 10!).

```{r quickAmp}
Test_Data_LONG <- amputeScaleScore(
                    ampute.data = data_to_ampute,
                    growth.config = growth_config_2021,
                    status.config = status_config_2021,
                    M=1)
```

We can inspect the results quickly to verify that we have a list with a single
(`data.table`) object

```{r quickCheck1}
  class(Test_Data_LONG)
  length(Test_Data_LONG)
  class(Test_Data_LONG[[1]])
  dim(Test_Data_LONG[[1]])
```

We will also make sure the proportion missing matches the `ampute.args$prop` value
default (0.3).

```{r quickCheck2}
Test_Data_LONG[[1]][YEAR == "2021" & VALID_CASE == "VALID_CASE",
                      list(NAs = round(sum(is.na(SCALE_SCORE))/.N, 3)),
                      keyby=list(CONTENT_AREA, GRADE)]
```

Note that the observed missingness is near the intended (30%). The
`mice::ampute.continuous` function does a decent job around 30% - 50% and worse
the further away you get from there! Corrections for this have been implemented,
but it's still a bit off. Depending on the desired precision, users will need to
experiment with `ampute.args$prop` argument values. The maximum proportion missing
is currently around 85%. Hopefully that extreme enough...


###  A more detailed amputation

Note that the `amputeScaleScore` does not return 2019 scores not used as priors.

```{r }
  table(Test_Data_LONG[[1]][, GRADE, YEAR])
```

We may want to add in 2019 priors (and we could also add in 2018 grades 6 to 8)
for some status and growth comparisons from 2019.  We will manually add those in
to the amputed data via the `additional.data` argument.

```{r addPriors}
  priors_to_add <- data_to_ampute[YEAR == "2019" & GRADE %in% c("7", "8")]
```

We also want to produce an interesting missingness pattern. For example, here we
will simulate missingness that is correlated with (low) achievement and economic
disadvantage (as indicated by Free/Reduced lunch status).  We will also assume
that these factors are compounded by the school level concentration of low
achievement and economic disadvantage.

```{r amp_vars}
  my.amp.vars <- c("SCHOOL_NUMBER", "SCALE_SCORE", "FREE_REDUCED_LUNCH_STATUS")
```

Specify the number of amputed data sets to create - 10 is the default.

```{r nAmps}
  MM <- 10
```

We will also add in the additional variable names that we want returned in the data
(`SCALE_SCORE_COMPLETE` and `ACH_LEV_COMPLETE`) in the `default.vars` argument.

```{r extendAmp}
#  Run 10 amputations with added priors
Test_Data_LONG <- amputeScaleScore(
                    ampute.data = data_to_ampute,
                    additional.data = priors_to_add,
                    growth.config = growth_config_2021,
                    status.config = status_config_2021,
                    M=MM,
                    default.vars = c("CONTENT_AREA", "GRADE",
                                     "SCALE_SCORE", "SCALE_SCORE_COMPLETE",
                                     "ACHIEVEMENT_LEVEL", "ACH_LEV_COMPLETE"),
                    demographics = c("FREE_REDUCED_LUNCH_STATUS", "ELL_STATUS",
                                     "IEP_STATUS", "ETHNICITY", "GENDER"),
                    institutions = c("SCHOOL_NUMBER", "DISTRICT_NUMBER"),
                    ampute.vars = my.amp.vars)
```

All 2019 prior scores are now added:

```{r extCheck1}
table(Test_Data_LONG[[3]][, GRADE, YEAR])

```

Inspect score missingness for all amputations. Look at both the raw number missing
in each grade and the proportions as well to verify.

```{r extCheck2}
  ##  Total missing scores in 2021 (note that we still have their "actual" score)
  table(Test_Data_LONG[[3]][is.na(SCALE_SCORE) & !is.na(SCALE_SCORE_COMPLETE), GRADE, YEAR])

  ##  Proportion missing scores in 2021 by GRADE
  for (m in seq(MM)) {
    print(Test_Data_LONG[[m]][YEAR == "2021" & VALID_CASE == "VALID_CASE",
            list(NAs = round(sum(is.na(SCALE_SCORE))/.N, 3)),
          keyby=list(CONTENT_AREA, GRADE)])
  }
```

# Missing data exploration/visualization

Now that we have some amputated data, we should evaluate it to make sure it looks
like what we wanted. In our assumptions, we expect to see more missing data for
students with lower 2019 scale scores, who are FRL, etc. The following visualization
from the `VIM` package are helpful in exploring. See the VIM package for more
context on these plots. In general, blue means observed data, red means missing.
A good introduction is given [here](https://www.datacamp.com/community/tutorials/visualize-data-vim-package).

```{r loadVIM}
  require(VIM)
```

Visualizations are easier with WIDE data, so we'll widen our datasets first.

```{r Amputed_long2wide}
  long.to.wide.vars <-
      c("GRADE", "SCALE_SCORE", "SCALE_SCORE_COMPLETE",
        "FREE_REDUCED_LUNCH_STATUS", "ELL_STATUS", "ETHNICITY")

  ###  Create a wide data set from each M amputated data sets
  Test_Data_WIDE <- vector(mode = "list", length = MM)

  for (m in seq(MM)) {
    Test_Data_WIDE[[m]] <- data.table::dcast(
      Test_Data_LONG[[m]][VALID_CASE == "VALID_CASE" & YEAR %in% c("2019", "2021")],
        ID + CONTENT_AREA ~ YEAR, sep=".", drop=FALSE, value.var=long.to.wide.vars)

    ##  Trim data of (107) cases with no data in 2019 or 2021
    Test_Data_WIDE[[m]] <- Test_Data_WIDE[[m]][!(is.na(GRADE.2019) & is.na(GRADE.2021))]

    ##  Fill in GRADE according to our expectations of normal progression
    Test_Data_WIDE[[m]][is.na(GRADE.2019), GRADE.2019 := as.numeric(GRADE.2021)-2L]
    Test_Data_WIDE[[m]][is.na(GRADE.2021), GRADE.2021 := as.numeric(GRADE.2019)+2L]

    ##  Exclude irrelevant GRADE levels
    Test_Data_WIDE[[m]] <- Test_Data_WIDE[[m]][GRADE.2021 %in% 3:8] # Current relevant GRADEs
    Test_Data_WIDE[[m]][!GRADE.2019 %in% 3:6, GRADE.2019 := NA] # Clean up prior GRADEs

    ##  Fill some more (but not all) - Demographics used in missing data plots ONLY
    Test_Data_WIDE[[m]][is.na(FREE_REDUCED_LUNCH_STATUS.2019), FREE_REDUCED_LUNCH_STATUS.2019 := FREE_REDUCED_LUNCH_STATUS.2021]
    Test_Data_WIDE[[m]][is.na(ELL_STATUS.2019), ELL_STATUS.2019 := ELL_STATUS.2021]
    Test_Data_WIDE[[m]][is.na(ETHNICITY.2019), ETHNICITY.2019 := ETHNICITY.2021]
  }
```

Start by looking at demographic variables.  Here iss missing in 2021 given 2019 `ETHNICITY`:

```{r vim1}
  spineMiss(as.data.frame(Test_Data_WIDE[[m]][, c("ETHNICITY.2019", "SCALE_SCORE.2021")]), interactive=FALSE)
```

There is not much differential impact, which makes sense given that `ETHNICITY`
was not used in the amputation calculation (but there may certainly be interactions
with other factors playing out here).

Next we'll look at the interplay between FRL, ELL and missingness

```{r vim2}
  mosaicMiss(
    Test_Data_WIDE[[m]][,
      c("FREE_REDUCED_LUNCH_STATUS.2019",
        "ELL_STATUS.2019",
        "SCALE_SCORE.2021")],
    highlight = 3, plotvars = 1:2, miss.labels = FALSE)
```

Here we see that FRL and ELL students are more likely to be missing in 2021. This
is very much the case for students who are both FRL and ELL. Too much so? We may
need to change the `ampute.var.weights` to downweight particular variables.

Now we'll look at missingness related to prior scale scores. First lets look at
what we started with in the `SGPdata::sgpData_LONG_COVID`.  We will widen it similar
to the way we widened the amputed data above.

```{r base_long2wide}
  sgpData_WIDE_COVID <- data.table::dcast(
    data_to_ampute[YEAR %in% c("2019", "2021") & VALID_CASE == "VALID_CASE"],
      ID + CONTENT_AREA ~ YEAR, sep=".", drop=FALSE, value.var=c("GRADE", "SCALE_SCORE"))

  ##    Trim things down with all NAs (79 cases) and fill in some missing info (partial.fill)
  sgpData_WIDE_COVID <- sgpData_WIDE_COVID[!(is.na(GRADE.2019) & is.na(GRADE.2021))]
  sgpData_WIDE_COVID[is.na(GRADE.2019), GRADE.2019 := as.numeric(GRADE.2021)-2L]
  sgpData_WIDE_COVID[is.na(GRADE.2021), GRADE.2021 := as.numeric(GRADE.2019)+2L]

  sgpData_WIDE_COVID <- sgpData_WIDE_COVID[GRADE.2021 %in% 5:8]
  sgpData_WIDE_COVID <- sgpData_WIDE_COVID[GRADE.2019 %in% 3:6]
```

Now we will create some plots from the unaltered data (note this is scores for all
grades, and scale is not standardized)

```{r vim3}
  histMiss(as.data.frame(sgpData_WIDE_COVID[, c("SCALE_SCORE.2019", "SCALE_SCORE.2021")]), breaks=25, interactive=FALSE)
  marginplot(as.data.frame(sgpData_WIDE_COVID[, c("SCALE_SCORE.2019", "SCALE_SCORE.2021")]))
  scattmatrixMiss(as.data.frame(sgpData_WIDE_COVID[, c("SCALE_SCORE.2019", "SCALE_SCORE.2021")]), interactive=FALSE)
```

Compare those with what we have in one of our amputated datasets: Change `m` to
see results from other amputations, and run the same plot types for unaltered/amputed
data back to back.

```{r vim4}
histMiss(as.data.frame(Test_Data_WIDE[[m]][, c("SCALE_SCORE.2019", "SCALE_SCORE.2021")]), breaks=25, interactive=FALSE, only.miss=FALSE)
marginplot(as.data.frame(Test_Data_WIDE[[m]][, c("SCALE_SCORE.2019", "SCALE_SCORE.2021")]))
scattmatrixMiss(as.data.frame(Test_Data_WIDE[[m]][, c("SCALE_SCORE.2019", "SCALE_SCORE.2021")]), interactive=FALSE)
```

What we see is that in the unaltered data the 2021 missing cases are more-or-less
randomly distributed given 2019 observed scores. After amputation, however, the
missing case distribution is heavily concentrated in lower prior score observations.

Check out similar plots for GRADE specific subsets:

```{r vim5}
histMiss(as.data.frame(Test_Data_WIDE[[m]][GRADE.2021 == "8",
  c("SCALE_SCORE.2019", "SCALE_SCORE.2021")]), breaks=25, interactive=FALSE)
marginplot(as.data.frame(Test_Data_WIDE[[m]][GRADE.2021 == "8",
  c("SCALE_SCORE.2019", "SCALE_SCORE.2021")]))

scattmatrixMiss(as.data.frame(Test_Data_WIDE[[m]][GRADE.2021 == "8",
    c("SCALE_SCORE.2019", "SCALE_SCORE.2021")]), interactive=FALSE)
```

Check status only grades for desired missing patterns:

```{r vim6}
mosaicMiss(Test_Data_WIDE[[m]][GRADE.2021 == "3",
    c("FREE_REDUCED_LUNCH_STATUS.2019", "ELL_STATUS.2019", "SCALE_SCORE.2021")],
  highlight = 3, plotvars = 1:2, miss.labels = FALSE)
```

Lastly, we can look at the amputed data relative to what we know to be "true".
Even though we didn't specify a MNAR missing pattern, we see that missingness is
highly correlated with unobserved (current) score because we've used prior scores
to determine the probability of missingness.  Given their high correlation we see
that play out in the amputed data as well.

```{r vim7}
histMiss(
  as.data.frame(Test_Data_WIDE[[m]][,
    c("SCALE_SCORE_COMPLETE.2021", "SCALE_SCORE.2021")]),
  breaks=25, interactive=FALSE)
histMiss(
  as.data.frame(Test_Data_WIDE[[m]][GRADE.2021 == "3",
    c("SCALE_SCORE_COMPLETE.2021", "SCALE_SCORE.2021")]),
  breaks=25, interactive=FALSE)
histMiss(
  as.data.frame(Test_Data_WIDE[[m]][GRADE.2021 == "8",
    c("SCALE_SCORE_COMPLETE.2021", "SCALE_SCORE.2021")]),
  breaks=25, interactive=FALSE)
```

The "marginplot" shows that the "observed" 2021 scores are identical to the "actual"
scores (blue dots along a perfect diagonal) - those are unaltered. The distribution
of the missing scores is skewed towards the lower end of achievement according to
the (red) boxplot.

```{r vim8}
histMiss(
  as.data.frame(Test_Data_WIDE[[m]][,
    c("SCALE_SCORE_COMPLETE.2021", "SCALE_SCORE.2021")]),
  breaks=25, interactive=FALSE)
marginplot(
  as.data.frame(Test_Data_WIDE[[m]][,
    c("SCALE_SCORE_COMPLETE.2021", "SCALE_SCORE.2021")]))

##    GROWTH grades only
histMiss(
  as.data.frame(Test_Data_WIDE[[m]][!GRADE.2021 %in% c("3", "4"),
    c("SCALE_SCORE_COMPLETE.2021", "SCALE_SCORE.2021")]),
  breaks=25, interactive=FALSE)
marginplot(
  as.data.frame(Test_Data_WIDE[[m]][!GRADE.2021 %in% c("3", "4"),
    c("SCALE_SCORE_COMPLETE.2021", "SCALE_SCORE.2021")]))

##    STATUS grades only
histMiss(
  as.data.frame(Test_Data_WIDE[[m]][GRADE.2021 %in% c("3", "4"),
    c("SCALE_SCORE_COMPLETE.2021", "SCALE_SCORE.2021")]),
  breaks=25, interactive=FALSE)
marginplot(
  as.data.frame(Test_Data_WIDE[[m]][GRADE.2021 %in% c("3", "4"),
    c("SCALE_SCORE_COMPLETE.2021", "SCALE_SCORE.2021")]))
```


# Missing data analysis

In the following we run through a quick pooling of the results from the 10 amputed
data sets to see how they differ from the "known truth". This analysis focuses on
school mean scale scores by grade, but similar analyses can be used with different
aggregates and/or using different variables of interest (e.g., SGPs) as well.

```{r pooledSchools}
school_means <- vector(mode = "list", length = length(Test_Data_LONG))
for (m in seq(MM)) {
  sch_mean <-
    Test_Data_LONG[[m]][YEAR %in% c("2019", "2021"), list(
        Mean_SS_SCHOOL = mean(SCALE_SCORE, na.rm=TRUE),
        SD_SS_SCHOOL = sd(SCALE_SCORE, na.rm=TRUE),
        Mean_SS_SCHOOL_COMPLETE = mean(SCALE_SCORE_COMPLETE, na.rm=TRUE),
        SD_SS_SCHOOL_COMPLETE = sd(SCALE_SCORE_COMPLETE, na.rm=TRUE),
        N = .N, PRESENT = sum(!is.na(SCALE_SCORE)),
        MISSING = sum(is.na(SCALE_SCORE))),
      keyby = c("YEAR", "CONTENT_AREA", "GRADE", "SCHOOL_NUMBER")]

  school_means[[m]] <- dcast(sch_mean, CONTENT_AREA + GRADE + SCHOOL_NUMBER ~ YEAR,
    value.var = c("Mean_SS_SCHOOL", "SD_SS_SCHOOL", "Mean_SS_SCHOOL_COMPLETE",
                  "SD_SS_SCHOOL_COMPLETE", "N", "PRESENT", "MISSING"),
    sep=".", drop=FALSE)
}

school_means <- rbindlist(school_means)
school_means[, c("Mean_SS_SCHOOL_COMPLETE.2019", "SD_SS_SCHOOL_COMPLETE.2019") := NULL]
setkeyv(school_means, c("CONTENT_AREA", "GRADE", "SCHOOL_NUMBER"))

pooled_school_means <- school_means[, list(
  Mean_SS_SCHOOL__2019 = mean(Mean_SS_SCHOOL.2019, na.rm=TRUE), # Should all be the same
  Mean_SD_SCHOOL__2019 = mean(SD_SS_SCHOOL.2019, na.rm=TRUE),
  Mean_SS_SCHOOL__2021 = mean(Mean_SS_SCHOOL.2021, na.rm=TRUE),
  SD_Mean_SS_SCHOOL__2021 = sd(Mean_SS_SCHOOL.2021, na.rm=TRUE),
  Mean_SD_SCHOOL__2021 = mean(SD_SS_SCHOOL.2021, na.rm=TRUE),
  Mean_SS_SCHOOL_COMPLETE__2021 = mean(Mean_SS_SCHOOL_COMPLETE.2021, na.rm=TRUE),
  Mean_SD_SCHOOL_COMPLETE__2021 = mean(SD_SS_SCHOOL_COMPLETE.2021, na.rm=TRUE),
  Mean_Present = mean(PRESENT.2021, na.rm=TRUE),
  Mean_Missing = mean(MISSING.2021, na.rm=TRUE)), keyby=c("CONTENT_AREA", "GRADE", "SCHOOL_NUMBER")]
```

Now we take a quick look at the pooled results. In the first table we see that
school mean scale scores have **increased** from 2019 to 2021. This is due in part
to the removal of more lower achieving students, on average.

```{r annualDiffs}
# Note this object has rows for each grade/content in all schools (grades 6-8 for elementary schools & 3-5 for middle schools)
  pooled_school_means[(!is.na(Mean_SS_SCHOOL__2019) & !is.na(Mean_SS_SCHOOL__2021)), # remove noted schools
    as.list(round(summary(Mean_SS_SCHOOL__2021-Mean_SS_SCHOOL__2019, na.rm=TRUE), 2)),
  keyby=c("CONTENT_AREA", "GRADE")]
```

We can also look at how different the 2021 amputed mean scores are from the "true"
mean scores. This shows us how biased the means are by the missing data (about 6
points too high, on average).

```{r truthDiffs}
  pooled_school_means[(!is.na(Mean_SS_SCHOOL_COMPLETE__2021) & !is.na(Mean_SS_SCHOOL__2021)),
    as.list(round(summary(Mean_SS_SCHOOL_COMPLETE__2021-Mean_SS_SCHOOL__2021, na.rm=TRUE), 2)),
  keyby=c("CONTENT_AREA", "GRADE")]
```

In addition to the artificially inflated mean scores, the "true" variation in
scores (within schools) is larger than in the amputed data.

```{r truthSD}
  pooled_school_means[(!is.na(Mean_SD_SCHOOL_COMPLETE__2021) & !is.na(Mean_SD_SCHOOL__2021)), # remove noted schools
    as.list(round(summary(Mean_SD_SCHOOL_COMPLETE__2021-Mean_SD_SCHOOL__2021, na.rm=TRUE), 2)),
  keyby=c("CONTENT_AREA", "GRADE")]
```
